Advancing Sustainable Mobility: A GIS-Based Multi-Criteria Decision Analysis for Equitable Electric Vehicle Supply Equipment Deployment in Philadelphia

### 2025-02-10 code/MCDA-EVSE.Rmd
interpret clusters

Third, at what scale should we implement local MCDA? In other words, what do we mean by "local"? This depends on the decision context and the spatial scale of the problem. There are many ways to solve the problem: census tract, neighborhoods, zip codes, etc. In our case, we chose a more systematic way to classify urban areas. We plan to first use PCA to reduce dimensionality by extracting key latent factors from the 9 variables. This also helps to remove redundancy (e.g., population density & driving population may be correlated). Then, we use k-means clustering to group similar grid cells into clusters based on the PCA-transformed data. This ensures that each cluster represents spatial areas with similar urban characteristics. Finally, we perform separate MCDA for each cluster, using different criteria weights. This allows us to capture local variations in EVSE suitability and make more informed decisions.

@@

@@

# combine with original fishnet data
fishnet_clustered <- left_join(fishnet_clustered, fishnet, by = "uniqueID")
```

@@

fishnet_clustered <- st_as_sf(fishnet_clustered)

@@

@@

# figure out common features in a cluster?
# bring in more variable information

@@

>>>>

Third, at what scale should we implement local MCDA? In other words, what do we mean by "local"? This depends on the decision context and the spatial scale of the problem. There are many ways to solve the problem: census tract, neighborhoods, zip codes, etc. In our case, we chose a more systematic way to classify urban areas. We plan to first use PCA to reduce dimensionality by extracting key latent factors from the 9 variables. This also helps to remove redundancy (e.g., population density & driving population may be correlated). Then, we use k-means clustering to group similar grid cells into clusters based on the PCA-transformed data. This ensures that each cluster represents spatial areas with similar urban characteristics. Finally, we perform separate MCDA for each cluster, using different criteria weights. This allows us to capture local variations in EVSE suitability and make more informed decisions. This is also called spectrum clustering, it clusters things better.

@@

```{r}
## plot top 20 loadings
top_k <- 5

## get pc1 and pc2
pc1 <- data.frame(loading = pca_fishnet$rotation[,1],
                  var = rownames(pca_fishnet$rotation),
                  pc = "PC1") 
pc2 <- data.frame(loading = pca_fishnet$rotation[,2],
                  var = rownames(pca_fishnet$rotation),
                  pc = "PC2")

# get top_k loadings of pc1 and pc2
pc1_top <- pc1 %>% arrange(-loading) %>% slice(1:top_k)
pc2_top <- pc2 %>% arrange(-loading) %>% slice(1:top_k)

rbind(pc1_top, pc2_top) %>%
  ggplot(aes(x = reorder(var, -loading), y = loading)) +
  geom_point() +
  ggtitle("Top loadings") +
  xlab("Gene") +
  facet_wrap(~pc, nrow = 1, scales = "free_x") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 1))
```

@@

# combine with original fishnet data
fishnet_clustered <- left_join(fishnet_clustered, fishnet_nowater, by = "uniqueID")
```

@@

#  4 clusters with 4 PCs
fishnet_clustered <- st_as_sf(fishnet_clustered)

@@

```{r cluster summary}

# figure out common features in a cluster?
# what are the commonalities within the cluster?

cluster_summary <- fishnet_clustered %>%
  group_by(Cluster) %>%
  dplyr::summarize(
    avg_density = mean(density, na.rm = TRUE),
    avg_driving_pop = mean(driving_pop, na.rm = TRUE),
    avg_evse_nn = mean(evse.nn, na.rm = TRUE),
    avg_TSFCA = mean(TSFCA, na.rm = TRUE),
    avg_elec_cnt = mean(elec_cnt, na.rm = TRUE),
    avg_parking_nn = mean(parking.nn, na.rm = TRUE),
    avg_police_fire = mean(policefire.nn, na.rm = TRUE),
    avg_zoning = mean(zone_score, na.rm = TRUE),
    avg_electric_use = mean(elect_use, na.rm = TRUE),
    count = n()  # Number of grid cells in each cluster
  )

cluster_summary 
```

From here, we can see that each cluster has different characteristics. For example, Cluster 1 has the highest average population density, driving population, registered EV, are more accessible to existing EVSE, and have more parking spaces nearby.

Cluster 4 also has high population density, and more than half of the population are in driving age. It has a fair number of registered electric vehicle, electric power usage is huge. Accessibility to the existing EVSE is low.

Cluster 3 has moderate population density, but also more than half of the population are in driving age. It has a lot of registered electric vehicle, but accessibility to EVSE is low.

Cluster 2 has low density and driving age population, but does have electric vehicle registration.

Looking at the map and based on our knowledge about Philadelphia, cluster 1 is likely to be Center City and University City, cluster 4 is likely to be Northeast Philadelphia and west Philadelphia, cluster 2 and 3 are both surban area, but cluster 3 is likely to be more affluent area while cluster 2 covers a lot of industrial area.

@@

ggplot(fishnet_clustered, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point() +
  labs(title = "Cluster Separation in PCA Space", x = "PC1", y = "PC2") +
  theme_bw() +
  theme(plot.subtitle = element_text(size = 12,face = "italic"),
        plot.title = element_text(size = 15, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=9), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, linewidth =0.8))
```

fishnet_clustered %>%
  st_drop_geometry() %>%
  dplyr::select(Cluster, driving_pop, evse.nn, elec_cnt, parking.nn) %>%
  mutate(
         evse.nn = scale_values(evse.nn),
         driving_pop = scale_values(driving_pop),
         elec_cnt = scale_values(elec_cnt),
         parking.nn= scale_values(parking.nn)
         ) %>% 
  pivot_longer(cols = -Cluster, names_to = "Variable", values_to = "Scaled_Value") %>% 
  ggplot(., aes(x = as.factor(Cluster), y = Scaled_Value, fill = as.factor(Cluster))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set3") +
  facet_wrap(~Variable, scales = "free_y") +  # Facet by variable
  labs(
    title = "Distribution of Scaled Criteria by Cluster",
    x = "Cluster",
    y = "Scaled Value (0-1)",
    fill = "Cluster"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    legend.position = "none",  # Remove legend since colors already represent clusters
    strip.text = element_text(size = 11, face = "bold")  # Styling facet labels
  )
```

Now we need to separate the fishnet data into 4 clusters based on the PCA-transformed data. We will then conduct MCDA for each cluster separately. This allows us to capture local variations in EVSE suitability and make more informed decisions.

```{r separate cluster}

# separate fishnet data into 4 clusters
cluster1 <- fishnet_clustered %>% filter(Cluster == 1)
cluster2 <- fishnet_clustered %>% filter(Cluster == 2)
cluster3 <- fishnet_clustered %>% filter(Cluster == 3)
cluster4 <- fishnet_clustered %>% filter(Cluster == 4)

@@

We will use the TOPSIS method to test out the local MCDA model

```{r construct performance table for local MCDA}

```{r local topsis operations}

# mutate(score = 0.12*scl_density + 0.18*scl_evse_re + 0.15*scl_elec + 0.1*scl_drivingpop + 0.1*scl_parking_re + 0.05*scl_policefire_re + 0.05*zone_score + 0.12*scl_acc_re +0.13*scl_elect_use) # self assign weight

# we need one set for each cluster
# reverse is min 
weights <- c(0.18, 0.12, 0.1, 0.12, 0.15, 0.1, 0.05, 0.05, 0.13)
criteriaMinMax <- c("min", "max", "max", "min", "max", "min", "min", "max", "max")
overall1 <- TOPSIS(performanceTable, weights, criteriaMinMax)
TOPSIS <- data.frame(uniqueID = names(overall1), values = overall1) %>%  
  mutate(uniqueID = as.integer(uniqueID))

we need to find a way to summarize the result of the local MCDA within each cluster.Probably the percentile rank of each grid cell within each cluster, compared to their average percentile rank in the entire city using global MCDA.
